---
title: "CL_HIV_Behavior_08"
author: "Group K"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Loading Libraries
```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
library(here)
library(ggplot2)

getwd()
list.files("data/raw_data")
```
#Load Dataset 
```{r}
hiv_df <- read_csv(here("raw_data", "hiv-behavior_national_zaf.csv"))
```

#Display Dataset content
```{r}
head(hiv_df)
```

#Remove the first row(meta data)
```{r}
hiv_df <- hiv_df[-1, ]
```
#dimensions 
```{r}
dim(hiv_df)
```
#Inspect Duplicated rows
```{r}
dup_check <- hiv_df %>%
  group_by(Indicator, SurveyYear, CharacteristicId, Value) %>%
  filter(n() > 1)

dup_check

```

```{r}
hiv_df <- hiv_df %>%
  distinct(Indicator, SurveyYear, CharacteristicId, Value, .keep_all = TRUE)

```
#Percentage Missing Values
```{r}

data.frame(
  Column = names(hiv_df),
  Missing_Percentage = paste0(round(colMeans(is.na(hiv_df)) * 100, 2), "%")
  )

```

```{r}
data.frame(
  Column = names(hiv_df),
  Missing_Data = paste0(colSums(is.na(hiv_df)))
  )
```

#check data types
```{r}
data.frame(
  Column = names(hiv_df),
  paste0(sapply(hiv_df, typeof))
)

```
#Check The structure of the dataset 
```{r}
str(hiv_df)

```
#Convert Data Types
```{r}
hiv_df <- hiv_df %>%
  mutate(
        Value = as.numeric(Value),
    Precision = as.numeric(Precision),
    SurveyYear = as.integer(SurveyYear),
    IndicatorOrder = as.integer(IndicatorOrder),
    CharacteristicId = as.integer(CharacteristicId),
    CharacteristicOrder = as.integer(CharacteristicOrder),
    IsTotal = as.logical(as.integer(IsTotal)),
    IsPreferred = as.logical(as.integer(IsPreferred)),
    SurveyYearLabel = as.integer(SurveyYearLabel),
    DenominatorWeighted = as.numeric(DenominatorWeighted),
    DenominatorUnweighted = as.numeric(DenominatorUnweighted),
  )
```
#check for unique values
```{r}
library(dplyr)
library(purrr)

# Summary table: column name, number of unique values, sample of unique values
n_sample <- 3

summary_tbl <- hiv_df %>%
  map_df(~ tibble(
    n_unique = n_distinct(.),
    sample_values = paste(head(unique(.), n_sample), collapse = ", ")
  ), .id = "column")


summary_tbl


```

#Drop the countries only onw unqiue value: reason, there is no useful information - county is also always za
```{r}
hiv_df <- hiv_df %>%

 select(
     -ISO3, 
    -DHS_CountryCode, 
    -CountryName, 
    -SurveyId,
    -ByVariableId, 
    -ByVariableLabel, 
    -IsTotal,
    -RegionId, 
    -SurveyYearLabel, 
    -SurveyType,
    -CharacteristicOrder
  )
```

#Assumed pattern, the missing values can be filled with the previous non missing value in the opposite attribute
```{r}


library(dplyr)
library(tidyr)

hiv_df <- hiv_df %>%
  mutate(
    DenominatorWeighted = if_else(DataId == "796612", 2955, DenominatorUnweighted),
    DenominatorWeighted = if_else(DataId == "795149", 2343, DenominatorUnweighted),
    DenominatorWeighted = if_else(DataId == "393814", 2842, DenominatorUnweighted),
    DenominatorWeighted = if_else(DataId == "407073", 1984, DenominatorUnweighted),
    DenominatorWeighted = if_else(DataId == "253304", 1235, DenominatorUnweighted),
    DenominatorWeighted = if_else(DataId == "253300", 848, DenominatorUnweighted) 
  )


hiv_df[
       c("DataId", "DenominatorWeighted", "DenominatorUnweighted")]

```

```{r}
ggplot(hiv_df, aes(x = DenominatorWeighted, y = DenominatorUnweighted)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  labs(title = "Scatterplot for Outlier Detection",
       x = "Denominator Weighted",
       y = "Denominator Unweighted") +
  theme_minimal()


ggplot(hiv_df, aes(y = DenominatorWeighted)) +
  geom_boxplot(fill = "lightblue", outlier.color = "red", outlier.shape = 16) +
  labs(title = "Boxplot of Denominator Weighted",
       y = "Denominator Weighted") +
  theme_minimal()

```

```{r}
dim(hiv_df)
```

#Outlier Handling
```{r}
# Calculate IQR boundaries
Q1_w <- quantile(hiv_df$DenominatorWeighted, 0.25, na.rm = TRUE)
Q3_w <- quantile(hiv_df$DenominatorWeighted, 0.75, na.rm = TRUE)
IQR_w <- Q3_w - Q1_w
lower_w <- Q1_w - 1.5 * IQR_w
upper_w <- Q3_w + 1.5 * IQR_w

Q1_uw <- quantile(hiv_df$DenominatorUnweighted, 0.25, na.rm = TRUE)
Q3_uw <- quantile(hiv_df$DenominatorUnweighted, 0.75, na.rm = TRUE)
IQR_uw <- Q3_uw - Q1_uw
lower_uw <- Q1_uw - 1.5 * IQR_uw
upper_uw <- Q3_uw + 1.5 * IQR_uw

# Cap values to the IQR limits
hiv_df <- hiv_df %>%
  mutate(
    DenominatorWeighted = pmin(pmax(DenominatorWeighted, lower_w), upper_w),
    DenominatorUnweighted = pmin(pmax(DenominatorUnweighted, lower_uw), upper_uw)
  )


```
#save cleaned data
```{r}
write_csv(hiv_df, here("cleaned_data", "hiv-behavior_cleaned.csv"))
```