---
littitle: "CL_Literacy_03"
author: "Group K"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Loading Libraries
```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
library(here)
library(purrr)

getwd()
list.files("data/raw_data")
```
#Load Dataset 
```{r}
lit_df <- read_csv(here("raw_data", "literacy_national_zaf.csv"))
```

#Disdplay Dataset content
```{r}
head(lit_df)
```

#Remove the first row(meta data)
```{r}
lit_df <- lit_df[-1, ]
```
#dimensions 
```{r}
dim(lit_df)
```

#Inspect Duplicated rows
```{r}
dup_check <- lit_df %>%
  group_by(Indicator, SurveyYear, CharacteristicId, Value) %>%
  filter(n() > 1)

dup_check

```

#perc na values
```{r}

data.frame(
  Column = names(lit_df),
  Missing_Percentage = paste0(round(colMeans(is.na(lit_df)) * 100, 2), "%")
  )

```

```{r}
data.frame(
  Column = names(lit_df),
  Missing_Data = paste0(colSums(is.na(lit_df)))
  )
```

#check data types
```{r}
data.frame(
  Column = names(lit_df),
  paste0(sapply(lit_df, typeof))
)

```
#Check The structure of the dataset 
```{r}
str(lit_df)

```

#Convert Data Types
```{r}
lit_df <- lit_df %>%
  mutate(
        Value = as.numeric(Value),
    Precision = as.numeric(Precision),
    SurveyYear = as.integer(SurveyYear),
    IndicatorOrder = as.integer(IndicatorOrder),
    CharacteristicId = as.integer(CharacteristicId),
    CharacteristicOrder = as.integer(CharacteristicOrder),
    IsTotal = as.logical(as.integer(IsTotal)),
    IsPreferred = as.logical(as.integer(IsPreferred)),
    SurveyYearLabel = as.integer(SurveyYearLabel),
    DenominatorWeighted = as.numeric(DenominatorWeighted),
    DenominatorUnweighted = as.numeric(DenominatorUnweighted),
  )
```
# Summary table: column name, number of unique values, sample of unique values
```{r}
library(purrr)
n_sample <- 3

summary_tbl <- lit_df %>%
  map_df(~ tibble(
    n_unique = n_distinct(.),
    sample_values = paste(head(unique(.), n_sample), collapse = ", ")
  ), .id = "column")


summary_tbl
```


#Drop the countries only onw unqiue value: reason, there is no useful information - county is also always za
```{r}
lit_df <- lit_df %>%
  select(
    -ISO3, 
    -DHS_CountryCode, 
    -CountryName, 
    -SurveyId,
    -CharacteristicOrder,
    -CharacteristicLabel,
    -ByVariableId, 
    -ByVariableLabel, 
    -IsTotal,
    -IsPreferred, 
    -RegionId, 
    -SurveyYearLabel, 
    -SurveyType
  )
```
#Another pattern found
```{r}


lit_df <- lit_df %>%
  fill(DenominatorWeighted, DenominatorUnweighted, .direction = "downup")

lit_df[
       c("DenominatorWeighted", "DenominatorUnweighted")]

```

```{r}
ggplot(acc_df, aes(x = Value, y = Value)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  labs(title = "Scatterplot for Outlier Detection",
       x = "Denominator Weighted",
       y = "Denominator Unweighted") +
  theme_minimal()

ggplot(acc_df, aes(y = Value)) +
  geom_boxplot(fill = "lightblue", outlier.color = "red", outlier.shape = 16) +
  labs(title = "Boxplot of Denominator Weighted",
       y = "Denominator Weighted") +
  theme_minimal()

```

#save cleaned data
```{r}
write_csv(lit_df, here("cleaned_data", "literacy_cleaned.csv"))
```