---
title: "CL_Maternal Mortality_04"
author: "Group K"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Loading Libraries
```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
library(here)
library(purrr)

getwd()
list.files("data/raw_data")
```
#Load Dataset 
```{r}
mam_df <- read_csv(here("raw_data", "maternal-mortality_national_zaf.csv"))
```

#Disdplay Dataset content
```{r}
head(mam_df)
```

#Remove the first row(meta data)
```{r}
mam_df <- mam_df[-1, ]
```
#dimensions 
```{r}
dim(mam_df)
```

#Inspect Duplicated rows
```{r}
dup_check <- mam_df %>%
  group_by(Indicator, SurveyYear, CharacteristicId, Value) %>%
  filter(n() > 1)

dup_check

```

#perc mising values
```{r}

data.frame(
  Column = names(mam_df),
  Missing_Percentage = paste0(round(colMeans(is.na(mam_df)) * 100, 2), "%")
  )

```

```{r}
data.frame(
  Column = names(mam_df),
  Missing_Data = paste0(colSums(is.na(mam_df)))
  )
```

#check data types
```{r}
data.frame(
  Column = names(mam_df),
  paste0(sapply(mam_df, typeof))
)

```
#Check The structure of the dataset 
```{r}
str(mam_df)

```

#Convert Data Types
```{r}
mam_df <- mam_df %>%
  mutate(
        Value = as.numeric(Value),
    Precision = as.numeric(Precision),
    SurveyYear = as.integer(SurveyYear),
    IndicatorOrder = as.integer(IndicatorOrder),
    CharacteristicId = as.integer(CharacteristicId),
    CharacteristicOrder = as.integer(CharacteristicOrder),
    IsTotal = as.logical(as.integer(IsTotal)),
    IsPreferred = as.logical(as.integer(IsPreferred)),
    SurveyYearLabel = as.integer(SurveyYearLabel),
    DenominatorWeighted = as.numeric(DenominatorWeighted),
    DenominatorUnweighted = as.numeric(DenominatorUnweighted),
  )
```
# Summary table: column name, number of unique values, sample of unique values
```{r}

n_sample <- 3

summary_tbl <- mam_df %>%
  map_df(~ tibble(
    n_unique = n_distinct(.),
    sample_values = paste(head(unique(.), n_sample), collapse = ", ")
  ), .id = "column")

summary_tbl
```
  

#Drop the countries only one unqiue value: reason, there is no useful information - county is also always za
```{r}
mam_df <- mam_df %>%
  select(
    -ISO3, 
    -DHS_CountryCode, 
    -CountryName, 
    -SurveyId,
    -ByVariableId, 
    -ByVariableLabel, 
    -IsTotal,
    -IsPreferred, 
    -RegionId, 
    -SurveyYearLabel, 
    -SurveyType
  )
```

```{r}
ggplot(mam_df, aes(x = Value, y = Value)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  labs(title = "Scatterplot for Outlier Detection",
       x = "Denominator Weighted",
       y = "Denominator Unweighted") +
  theme_minimal()


ggplot(mam_df, aes(y = Value)) +
  geom_boxplot(fill = "lightblue", outlier.color = "red", outlier.shape = 16) +
  labs(title = "Boxplot of Value",
       y = "Denominator Weighted") +
  theme_minimal()


```

#save cleaned data
```{r}
write_csv(acc_df, here("cleaned_data", "maternal-mortality_cleaned.csv"))
```
